MetaTracts - A Method for Robust Extraction and Visualization of Carbon Fiber Bundles in Fiber Reinforced Composites

### 2014-09-25 chapters/MetaTracts.tex
Updated final ppt with the comparison image

\begin{figure}[tb]
\centering
%\includegraphics[width=0.45\textwidth]{imagesMT2014/MT_data_4_2.png}
\includegraphics[width=0.5\textwidth, trim = 0mm 110mm 0mm 0mm, clip,]{images_pvis/figure1}
\caption{Data characteristics: (a) Rendering of the data. (b) 2D slices along the X-axis and (c) 2D slice along the Z-axis. (d) Magnified region marked in green square (c). Multiple fiber bundles cross and are indistinguishable by visual inspection alone. }
\label{fig:data-char}
\end{figure}
\section {Data Characteristic and Assumptions}
\label{sec:char_data}
Figure \ref{fig:data-char} shows dataset 1 with woven fiber bundles. Dataset 1 is 450$\times$300$\times$500 voxels in size with isotropic resolution of 2$\mu m$. The data type is unsigned 8 bit integer. Figure~\ref{fig:data-char} clearly shows the weaving structure of the composite unit cell as regional subset of the recurring fiber bundle weaving pattern. The unit cell is used as a basis structure for CFRP manufacturing. The alignment and weaving pattern of the fiber bundles in the cloth influences the strength and stiffness properties of the resulting material. Figure~\ref{fig:data-char}a shows a volume rendering of the dataset. Figure~\ref{fig:data-char}b shows 2D slices along the $X$- and Figure~\ref{fig:data-char}c in $Z$-axis. Figure~\ref{fig:data-char}d shows the magnified version of the green region of interest. 
The resolution of dataset 1 renders the individual carbon fibers in a fiber bundle hard to resolve using 2D slice images only. Figure~\ref{fig:data-char}d contains two bundles going in opposite directions. The separation between two fiber bundles is barely visible. The carbon fibers themselves do not differ much from the underlying epoxy matrix in terms of attenuation which poses an additional challenge for characterization. The fiber bundles may differ in terms of the amount of fibers in the bundle. Figure~\ref{fig:data-char}a shows the large variation in cross section sizes among the bundles.
Depending on the weaving pattern, the fiber bundles cross each other in different orientations. In general, the number of orientations is limited by the weaving pattern. Nevertheless, the weaving pattern may cause the individual bundles to be curved. In consequence individual fibers may be adjacent in Euclidean distance but belong to different bundles.

@@

Fig. \ref{fig:flowchart} shows our pipeline. MetaTracts generations  is explained in Section.~\ref{sec:approach} and  finding plausible fiber bundles in Section~\ref{subsec:fiber-bundles}, Section~\ref{sec:vis} talks about visualization extensions.
%\subsection {Extracting MetaTracts}
%\label{subsec:ExFiberCyl}
%We explain the different steps of MetaTracts extraction.

@@

The input to the first stage of our pipelines, is the original scalar dataset in a uniform lattice grid in $\mathbb{R}^3$. As output, each grid location becomes associated with a vector representing the local orientation and a real value [0,1] which represents a measure of the reliability of the local orientation. We approximate the local orientation by eigenvalue analysis of the Hessian matrix computed at each voxel. The principal directions, in which the local second order structure of the image can be decomposed, are given by the eigen decomposition of the Hessian matrix which describes the local curvature.
The eigenvector corresponding to the smallest eigenvalue gives the direction along which the curvature is smallest. This direction coincides with the direction of the tubular structure.

Frangi et al.~\cite{Frangi1998} introduced a process that searches for geometric structures which are tubular. They define a measure based on two geometric ratios of the second order ellipsoid given by the local Hessian matrix to measure the ``vesselness'' criterion.  In order to determine reliable Hessians, we compute a similar metric. We include their work here for completeness and direct the reader to ~\cite{Frangi1998} for details. Let $\lambda_{K}$ be the eigenvalue with the $K^{th}$ smallest magnitude. Here $|{\lambda}_{1}| \leq| {\lambda}_{2}|\leq| {\lambda}_{3}| $ are the eigenvalues of the Hessian matrix. Specifically, a pixel belonging to a vessel region will have small $\lambda_{1}$ ($|\lambda_{1}|\approx 0$) and $\lambda_{2}$, $\lambda_{3}$ of large magnitude and of equal sign($|\lambda_{1}| \ll |\lambda_{2}|$ and $|\lambda_{2}|\approx |\lambda_{3}|$). The sign indicates if the vessel is bright in a dark background or dark in a bright background. In our case the individual fibers are bright ($\lambda_2,\lambda_3 < 0$). The following measures are defined in their work~\cite{Frangi1998}.  

@@

In equation~\ref{RB}, $\mathcal{R_{B}}$ provides a measure of deviation from a blob like structure while in equation~\ref{RA}, $\mathcal{R_{A}}$ distinguishes between plate-like and line-like structure. Grey-scale variations and close proximity of the fibers in our data make the Hessians computed at each voxels susceptible to errors. Thus we compute the ``vesselness'' measure to determine which locations in the volume provide reliable local orientation.

@@

First, in biology vesselness is computed for different scales because the vessels can be of different sizes. In our case, usually the width of individual fibers are known a priori.

@@

In Figure~\ref{fig:reliable_hessian} we see the intermediate results of the local orientation computation only at locations where $R_{H}$ is greater than the cutoff threshold. The unit vector representing the principal direction has been mapped to RGB color space. Figure~\ref{fig:reliable_hessian}a shows the entire data set. Regions where the principal direction is parallel to the X axis are red in color, directions parallel to Y are green and those parallel to the Z are blue. Figure~\ref{fig:reliable_hessian}b,c show 2D slices along the Z and X axis respectively. Figure~\ref{fig:reliable_hessian}d shows a magnified region of interest. Note, the dark regions within a bundles are regions where the $R_H$ is less than the threshold and are unreliable. The bundles are not uniformly colored, the Hessians and the corresponding directions are noisy.

@@

We note some intrinsic differences between the DTI and our XCT data. Fiber traces can be created in DTI using a standard fiber tracking algorithm following the principal direction of diffusion using a fourth order Runge-Kutta method~\cite{Brun2003}. The principal direction based on the Hessian matrix works best when the tubular structures are well separated from the surrounding, this is not the case for our data. The local orientaion at each voxel is inherently more noisy. Thus techniques popular in DTI cannot be directly applied to XCT data.

%
\begin{figure}[tb]
\centering
%\includegraphics[width=0.45\textwidth]{images_pvis/figure3}
\includegraphics[width=0.5\textwidth,  trim = 0mm 90mm 0mm 0mm, clip]{images_pvis/figure3}
\caption{Reliable Hessians. (A) Colored according to Orientation vector mapped to RGB. (B) 2D slice along Z and (C) along X axis. (D) Magnified region marked in green square (in C).}
\label{fig:reliable_hessian}
\end{figure}

\subsection {MetaTracts properties and description} 
\label{subsec:mtprop}
%\begin{figure*}
%  \centering
%  \subfloat[Meta tract generation in 2D]{\includegraphics[width=0.5\linewidth]{imagesMetaTracts/algo-1b.eps}}
%  \subfloat[Orientation based similarity measure]{\includegraphics[width=0.5\linewidth]{imagesMetaTracts/algo-1c.eps}}
%  \caption{Simplified examples of meta tract generation process and orientation based similarity measure}
%  \label{fig:algo}
%\end{figure*}
%\begin{figure*}
%  \centering
%  \subfloat[]{\includegraphics[width=0.2\textwidth]{imagesMetaTracts/algo-1d}}
%  %\subfloat[Orientation based similarity measure]{\includegraphics[width=0.5\linewidth]{imagesMetaTracts/algo-1c.eps}}
%  \caption{Simplified example of meta tract generation process.}
%  \label{fig:algo}
%\end{figure*}
The input to this step is a grid in $\mathbb{R}^3$ where each grid point is associated with a normalized local orientation and a corresponding $R_H$ value.
Traditional integral curve based techniques cannot be used to extract fiber bundle traces from the reliable Hessians because of the spurious nature of the Hessian based local orientation. Thus instead of building fiber traces, we wish to find an abstract representation of the fibers. We build on the key assumptions on the data, namely `local orientation' and `connectivity' while taking into account the noise and lack of resolution. We do this by interpreting the underlying geometric structure of the fibers as a set of polylines (cylinders). MetaTacts  are a coarse and simple approximation of integral curves in the form of a continuous chain of cylindrical tubes in $\mathbb{R}^3$ traversing the fiber bundles embedded in the input data. MetaTracts share the following \textit{properties}:

\begin{enumerate}[noitemsep]
\item{Each MetaTract is associated with a continuous set of cylinders.}
\item{Each MetaTract is associated with a start point which is a grid vertex.}
\item{Each individual cylinder in a MetaTract has a constant length, a constant radius and a start point (which is also a grid vertex).}
\item{Each cylinder in a MetaTract (except for the first one) is connected to the previous cylinder at its start point.}
\item{Each cylinder is locally parallel to the local orientation at it's start point.}
\end{enumerate}

@@

\subsection {MetaTracts generation}
\label{subsec:MetaTracts-generation}
 In this section we explain the process of MetaTracts generation in $\mathbb{R}^2$. The procedure trivially extends to $\mathbb{R}^3$. In $\mathbb{R}^2$ all the above properties hold except that cylinders are replaced by rectangles. The reliable Hessian step  generates a local orientation in $\mathbb{R}^2$ and a $R_H$[0,1]. In Figure \ref{fig:algo} all regions which have reliable Hessian less than the threshold have been marked in blue. Figure~\ref{fig:algo}c shows a single cylinder of a MetaTracts. Let the seed point associated with the MetaTract be grid point $C_{p}$ (property 2 in Section~\ref{subsec:mtprop}). The local orientation at $C_p$ as computed in the step above is $N_P$ and is given by the dark green arrow. We generate a rectangle of length $L$ and radius $R$, the green dashed rectangle shows the rectangle generated at $C_p$ (properties 3 and 5 in Section~\ref{subsec:mtprop}).

@@

\begin{itemize}
\item \textit{Orientation similarity}: We want the orientation of the start points for the consecutive cylinders to have a similar orientations ($N_p$). 
\item \textit{Large distance}: We want the MetaTracts to traverse the data using as few cylinders as possible. Thus, we want the distance between $C_p$ and the start point for the next cylinder to be as large as possible. We measure distance of a candidate vertex from $C_p$ by projecting the Euclidean distance between them onto $N_p$. For example the Euclidean distance between the blue and the orange vertices projected onto $N_P$ gives distance (Figure~\ref{fig:algo}B). We refer to this perpendicular projection distance as \textit{`projected\_dist'}.

\end{itemize}
% \begin{equation}\label{eqn:algo_1}
% d=\frac{1}{3}e^{(\frac{-angle^{2}}{\alpha^{2}})}+\frac{2}{3}(1-e^{\frac{-{PerpDist}^{2}}{\beta^{2}}})
% \end{equation}
We define a priority for each of the candidate vertices based on the above factors.
For each cylinder in a MetaTract, we put the \textit{candidate vertices} of it in a priority queue based on Equation \ref{eqn:algo_1}.
\begin{equation}
Priority = \gamma_1 e^{(-angle^2 / \alpha^2)} + \gamma_2e^{(-projected\_dist^2 / \beta^2)}
\label{eqn:algo_1}
\end{equation}
$\gamma_1$, $\gamma_2$ are the  weights ($\mathbb{R}_{\ge 0}$)  which decide how the priority depends on the affine combination of the two factors. For all our cases we use $\gamma_1$ = $1 / 3 $ and $\gamma_2$ = $2 / 3$. In general we suggest $\gamma_1 +\gamma_2 = 1 $ and $\gamma_1 \leq \gamma_2$. At each iteration we pick the top element in the priority queue generate the corresponding cylinder and repeat the steps. Essentially, Equation~\ref{eqn:algo_1} selects a grid point which is the furthest from the current point and going in a similar direction. This gives it the advantage of tackling noise/errors in local orientation better than integral curves by looking at multiple choices for vertices and avoiding intra cell interpolation in an already noisy environment.

In this particular example in Figure~\ref{fig:algo}b we select the orange grid vertex next and repeat the process. We do not select the purple vertex (Figure~\ref{fig:algo}c) because it is not a candidate vertex. If we generate tracts that have erroneous local orientations they will not be able to find further possible candidate vertices and will be of short length. Short tracts are then removed. The MetaTracts generated are shown in Figure~\ref{fig:length_distribution}a. The MetaTracts are colored with the mean orientation direction mapped to the RGB space. Figure\ref{fig:length_distribution}a shows a single slice of YZ plane. That consistent orientation is a key intrinsic feature in the data becomes visually pronounced.
We apply uniform, dense seeding to the XCT volume data to trace and generate the tracts of fiber bundles. Figure~\ref{fig:length_distribution}b shows the MetaTracts of a particular bundle colored according to their individual length.

@@

The output of the above step is a set of MetaTracts. The next step is to cluster the MetaTracts into bundles. 
Carbon fibers have both orientation and geometric proximity information, both of which can be used for clustering. Experimentally we found that orientation measure was very reliable while measures of geometric proximity had problems with fibers which partially overlap (Figure~\ref{fig:length_distribution}b).  
Instead of creating a heuristic which artificially combines the orientation and geometric proximity measures, we first cluster based on orientation and then further subdivide each orientation cluster using geometric proximity.
We found that different clustering techniques performed preferably for different measures.  The two step clustering also made the problem more tractable and  the parameters intuitive for the end users. For the orientation clustering, we use dimension reduction followed by K-means clustering. We then use hierarchical clustering to further subdivide each cluster based on the geometric proximity measures.

@@

\subsection {Orientation based clustering}
%We use the inherent features expressed as weaving pattern of our data to pre-cluster the metaTracts into the number of classes based on the major directions of the pattern.
%It is important to note that this step broadly divides the MetaTracts into classes based on the major orientations and not individual fiber bundles. 
This step divides the individual MetaTracts into classes based on their major orientations. In order to cluster MetaTracts going in same directions, we use a spectral embedding technique called Laplacian eigenmaps which was originally introduced by Belkin and Niyogi \cite{Belkin01} and later used in DTI fiber coloring~\cite{Brun2003} and fiber clustering~\cite{Brun2004}. The key notion is to find a suitable similarity measure to define the weights of an undirected weighted graph. The nodes being the data points and the edges being the weights. Then, an eigenvalue problem is solved, which maps the manifold embedded in the graph into a lower dimensional space while preserving the graph structure.
This lower dimensional space can then be clustered more efficiently than the higher dimensional space. 
Let $G$ be the graph, we compute the eigenvalues and eigenvectors for the generalized eigenvector problem:$L\textit{\textbf{f}}=\lambda D\textit{\textbf{f}}$,
%\begin{equation}\label{equn:eigenMaps}
%L\textbf{f}=\lambda D\textbf{f}
%\end{equation}
where $D$ is the diagonal weight matrix and $L$ is the Laplacian matrix. The eigenvector \textbf{${f}_{0}$} corresponding to the eigenvalue 0 is left out and the next $m$, {\textbf{${f}_{1}$} through \textbf{${f}_{m}$}} eigenvectors are used to embed in an $m$-dimensional space. K-means clustering is then used to cluster in this lower dimensional space.

Brun et al.~\cite{Brun2003} based their measure on the simple assumption that two traces with similar endpoints should be considered similar. Brun et al. ~\cite{Brun2004} based their measure on a 9-D fiber descriptors. In our case, each MetaTract is a data point. We introduce a simple \textit{orientation based similarity measure}. There is no spatial information involved, just a partition of the MetaTracts based on orientation which is already inherent in the data.
Following this, given a pair of  MetaTracts, we define the edge weights between two tracts as the cosine of the maximum angle between the local orientations ($N_P$) of all pairs of start points ($C_P$) between the two MetaTracts. The edge weights give a distance matrix representing the distance between each pair of points. Using the Belkin and Niyogi algorithm we ``embed" these points in a low dimensional space where the Euclidean distance between points approximates the distance between points given by the distance matrix. 
%
%August 13 2014
%For all our tests the fibers were then embedded in a $\eta$ low dimensional space. The number of dimensions in the high dimensional space is equal to the number of datapoints or the number of MetaTracts in our case.
%For all our test cases we set $\eta$ to be five (see sec~\ref{sec:param_choices} for parameter choices for $\eta$.).
%  
We used conventional K-means for clustering this lower dimensional space. Here $K$ is supplied by domain knowledge of the number of major fiber bundle directions of the woven structure.
For our test case there are two major directions of the fiber bundles and $K$ was set to 2. Due to the advantages offered by the dimensionality reduction, even if there are curved fibers bundles the user has to provide just the major fiber bundle directions of the weaving pattern. Figure~\ref{fig:orientation_clustering}b shows the result of the K-means clustering with the data points projected to the top three eigenvectors as the major axes. As is expected there is a clear distinction based on fiber orientation. Figure~\ref{fig:orientation_clustering}a,c shows the MetaTracts colored according to clustering results.

@@

Unfortunately, the Hausdorff distance does not work well for our application since a fiber bundle may have many MetaTract which only partially covers the bundle (Figure~\ref{fig:length_distribution}b). If a MetaTract $P$ covers only part of the fiber bundle covered by $Q$, then $H_{dir}(P,Q)$ will be very small while $H_{dir}(Q,P)$ will be large.

@@

In comparison Fig.~\ref{fig:comparison} shows the MetaTracts in Fig.~\ref{fig:length_distribution}(A) being hierarchically clustered directly using proximity alone into 10,15 and 20 clusters (ground truth being 11). As discussed above, single linkage hierarchical clustering in the presence of overlapping fibers tends to create large clusters and small, low cardinality outliers. In Fig.~\ref{fig:comparison} when ``h" is 10 two large multi-bundle incorrect cluster is generated and the rest are outliers. As ``h" increases, some appropriate bundles start to form. Even at ``h" equals 20, the top bundles all clusters together into one large incorrectly identified bundle.

>>>>

\begin{figure}[tb]
\centering
%\includegraphics[width=0.5\textwidth, trim = 0mm 110mm 0mm 0mm, clip,]{images_pvis/figure1}
\includegraphics[width=0.5\textwidth]{images_pvis/data-char.pdf}
\caption{Data characteristics: (a) Rendering of the data. (b) 2D slices along the X-axis and (c) 2D slice along the Z-axis. (d) Magnified region marked in green square (c). Multiple fiber bundles cross and are indistinguishable by visual inspection alone. }
\label{fig:data-char}
\end{figure}
\section {Data Characteristic and Assumptions}
\label{sec:char_data}
Figure \ref{fig:data-char} shows dataset 1 with woven fiber bundles. Dataset 1 is 450$\times$300$\times$500 voxels in size with isotropic resolution of 2$\mu m$. The data type is unsigned 8 bit integer. Figure~\ref{fig:data-char} clearly shows the weaving structure of the composite unit cell as regional subset of the recurring fiber bundle weaving pattern. The unit cell is used as a basis structure for CFRP manufacturing. The alignment and weaving pattern of the fiber bundles in the cloth influences the strength and stiffness properties of the resulting material. Figure~\ref{fig:data-char}a shows a volume rendering of the dataset. Figure~\ref{fig:data-char}b shows 2D slices along the $X$- and Figure~\ref{fig:data-char}c in $Z$-axis. Figure~\ref{fig:data-char}d shows the magnified version of the green region of interest. 
The resolution of dataset 1 renders the individual carbon fibers in a fiber bundle hard to resolve using 2D slice images only. Figure~\ref{fig:data-char}d contains two bundles going in opposite directions. The separation between two fiber bundles is barely visible. The carbon fibers themselves do not differ much from the underlying epoxy matrix in terms of attenuation which poses an additional challenge for characterization. The fiber bundles may differ in terms of the amount of fibers in the bundle. Figure~\ref{fig:data-char}a shows the large variation in cross section sizes among the bundles.
Depending on the weaving pattern, the fiber bundles cross each other in different orientations. In general, the number of orientations is limited by the weaving pattern. Nevertheless, the weaving pattern may cause the individual bundles to be curved. In consequence individual fibers may be adjacent in Euclidean distance but belong to different bundles.

@@

Fig. \ref{fig:flowchart} shows our pipeline. MetaTracts generations  is explained in Section.~\ref{sec:approach} and  finding plausible fiber bundles in Section~\ref{subsec:fiber-bundles}, Section~\ref{sec:vis} talks about visualization extensions.

Extracting the MetaTracts consists of two  major steps. First we extract a local orientation vector at each grid location (section~\ref{subsec:rh}). Second, we compute coarse poly-cylinders called MetaTracts which traverse the data (section~\ref{subsec:mtprop}).

%\subsection {Extracting MetaTracts}
%\label{subsec:ExFiberCyl}
%We explain the different steps of MetaTracts extraction.

@@

The input to the first stage of our pipeline, is the original scalar dataset in a uniform lattice grid in $\mathbb{R}^3$. As output, each grid location becomes associated with a vector representing the local orientation and a real value [0,1] that represents a measure of the reliability of the local orientation. We approximate the local orientation by eigenvalue analysis of the Hessian matrix computed at each voxel. The principal directions, in which the local second order structure of the image can be decomposed, are given by the eigen decomposition of the Hessian matrix which describes the local curvature.
The eigenvector corresponding to the smallest eigenvalue gives the direction along which the curvature is smallest. This direction coincides with the direction of the tubular structure.

Frangi et al.~\cite{Frangi1998} introduced a process that searches for geometric structures which are tubular. They define a measure based on two geometric ratios of the second order ellipsoid given by the local Hessian matrix to measure the ``vesselness'' criterion.  In order to determine reliable Hessians, we compute a similar metric. We include their work here for completeness and direct the reader to ~\cite{Frangi1998} for details. Let $\lambda_{K}$ be the eigenvalue with the $K^{th}$ smallest magnitude. Here, $|{\lambda}_{1}| \leq| {\lambda}_{2}|\leq| {\lambda}_{3}| $ are the eigenvalues of the Hessian matrix. Specifically, a pixel belonging to a vessel region will have small $\lambda_{1}$ ($|\lambda_{1}|\approx 0$) and $\lambda_{2}$, $\lambda_{3}$ of large magnitude and of equal sign($|\lambda_{1}| \ll |\lambda_{2}|$ and $|\lambda_{2}|\approx |\lambda_{3}|$). The sign indicates if the vessel is bright in a dark background or dark in a bright background. In our case the individual fibers are bright ($\lambda_2,\lambda_3 < 0$). The following measures are defined in~\cite{Frangi1998}.  

@@

In Equation~\ref{RB}, $\mathcal{R_{B}}$ provides a measure of deviation from a blob like structure while in Equation~\ref{RA}, $\mathcal{R_{A}}$ distinguishes between plate-like and line-like structure. Grey-scale variations and close proximity of the fibers in our data make the Hessians computed at each voxels susceptible to errors. Thus, we compute the ``vesselness'' measure to determine which locations in the volume provide reliable local orientation.

@@

First, vesselness in biology is computed for different scales because the vessels can be of different sizes. In our case, usually the width of individual fibers are known a priori.

@@

In Figure~\ref{fig:reliable_hessian} we see the intermediate results of the local orientation computation only at locations where $R_{H}$ is greater than the cutoff threshold. The unit vector representing the principal direction has been mapped to RGB color space. Figure~\ref{fig:reliable_hessian}a shows the entire data set. Regions where the principal direction is parallel to the X axis are red in color, directions parallel to Y are green and those parallel to the Z are blue. Figure~\ref{fig:reliable_hessian}b,c show 2D slices along the Z and X axes respectively. Figure~\ref{fig:reliable_hessian}d shows a magnified region of interest. Note, the dark regions within a bundles are regions where the $R_H$ is less than the threshold and are unreliable. The bundles are not uniformly colored, the Hessians and the corresponding directions are noisy.

@@

We note some intrinsic differences between the DTI and our XCT data. Fiber traces can be created in DTI using a standard fiber tracking algorithm following the principal direction of diffusion using a fourth order Runge-Kutta method~\cite{Brun2003}. The principal direction based on the Hessian matrix works best when the tubular structures are well separated from the surrounding. This is not the case for our data. The local orientaion at each voxel is inherently more noisy. Thus, techniques popular in DTI cannot be directly applied to XCT data.

%
\begin{figure}[tb]
\centering
%\includegraphics[width=0.45\textwidth]{images_pvis/figure3}
%\includegraphics[width=0.5\textwidth,  trim = 0mm 90mm 0mm 0mm, clip]{images_pvis/figure3}
\includegraphics[width=0.5\textwidth]{images_pvis/reliable_hessian.pdf}
\caption{Reliable Hessians. (a) Colored according to Orientation vector mapped to RGB. (b) 2D slice along Z and (c) along X axis. (d) Magnified region marked in green square (in c).}
\label{fig:reliable_hessian}
\end{figure}

\subsection {MetaTracts properties and description} 
\label{subsec:mtprop}
%\begin{figure*}
%  \centering
%  \subfloat[Meta tract generation in 2D]{\includegraphics[width=0.5\linewidth]{imagesMetaTracts/algo-1b.eps}}
%  \subfloat[Orientation based similarity measure]{\includegraphics[width=0.5\linewidth]{imagesMetaTracts/algo-1c.eps}}
%  \caption{Simplified examples of meta tract generation process and orientation based similarity measure}
%  \label{fig:algo}
%\end{figure*}
%\begin{figure*}
%  \centering
%  \subfloat[]{\includegraphics[width=0.2\textwidth]{imagesMetaTracts/algo-1d}}
%  %\subfloat[Orientation based similarity measure]{\includegraphics[width=0.5\linewidth]{imagesMetaTracts/algo-1c.eps}}
%  \caption{Simplified example of meta tract generation process.}
%  \label{fig:algo}
%\end{figure*}
The input to this step is a grid in $\mathbb{R}^3$ where each grid point is associated with a normalized local orientation and a corresponding $R_H$ value.
Traditional integral curve based techniques cannot be used to extract fiber bundle traces from the reliable Hessians because of the spurious nature of the Hessian based local orientation. Thus instead of building fiber traces, we wish to find an abstract representation of the fibers. We build on the key assumptions on the data, namely ``local orientation" and ``connectivity" while taking into account the noise and lack of resolution. We do this by interpreting the underlying geometric structure of the fibers as a set of polylines (cylinders). MetaTacts  are a coarse and simple approximation of integral curves in the form of a continuous chain of cylindrical tubes in $\mathbb{R}^3$ traversing the fiber bundles embedded in the input data. MetaTracts share the following \textit{properties}:

\begin{enumerate}[noitemsep]
\item{Each MetaTract is associated with a continuous set of cylinders.}
\item{Each MetaTract is associated with a start point which is a grid vertex.}
\item{Each individual cylinder in a MetaTract has a constant length, a constant radius and a start point (which is also a grid vertex).}
\item{Each cylinder in a MetaTract (except for the first one) is connected to the previous cylinder at its start point.}
\item{Each cylinder is locally parallel to the local orientation at its start point.}
\end{enumerate}

@@

\subsection {MetaTracts generation}
\label{subsec:MetaTracts-generation}
 In this section we explain the process of MetaTracts generation in $\mathbb{R}^2$.The procedure trivially extends to $\mathbb{R}^3$. In $\mathbb{R}^2$, all the above properties hold except that cylinders are replaced by rectangles. The reliable Hessian step  generates a local orientation in $\mathbb{R}^2$ and a reliability measure $R_H$[0,1]. In Figure \ref{fig:algo}, all regions which have unreliable Hessian have been marked in blue. Figure~\ref{fig:algo}c shows a single cylinder of a MetaTracts. Let the seed point associated with the MetaTract be grid point $C_{p}$ (property 2 in Section~\ref{subsec:mtprop}). The local orientation at $C_p$ as computed in the step above is $N_P$ and is given by the dark green arrow. We generate a rectangle of length $L$ and radius $R$, the green dashed rectangle shows the rectangle generated at $C_p$ (properties 3 and 5 in Section~\ref{subsec:mtprop}).

@@

\begin{itemize}
\item \textit{Orientation similarity}: The orientation of the start points ($N_p$) for the consecutive cylinders should be similar. 
\item \textit{Large distance}: The MetaTracts should traverse the data using as few cylinders as possible. Thus, the distance between $C_p$ and the start point for the next cylinder should be as large as possible. We measure the distance of a candidate vertex from $C_p$ by projecting the Euclidean distance between them onto $N_p$.

For example, in Figure~\ref{fig:algo}b the distance is measured as the Euclidean distance between the purple and the orange vertices projected on to $N_p$. We refer to this perpendicular projection distance as \textit{`projected\_dist'}.

\end{itemize}
% \begin{equation}\label{eqn:algo_1}
% d=\frac{1}{3}e^{(\frac{-angle^{2}}{\alpha^{2}})}+\frac{2}{3}(1-e^{\frac{-{PerpDist}^{2}}{\beta^{2}}})
% \end{equation}
We define a priority for each of the candidate vertices based on the above factors.
For each cylinder in a MetaTract, we put its \textit{candidate vertices} in a priority queue based on Equation \ref{eqn:algo_1}.
\begin{equation}
Priority = \gamma_1 e^{(-angle^2 / \alpha^2)} + \gamma_2e^{(-projected\_dist^2 / \beta^2)}
\label{eqn:algo_1}
\end{equation}
$\gamma_1$, $\gamma_2$ are the  weights ($\mathbb{R}_{\ge 0}$)  which decide how the priority depends on the affine combination of the two factors. For all our cases, we use $\gamma_1$ = $1 / 3 $ and $\gamma_2$ = $2 / 3$. In general, we suggest $\gamma_1 +\gamma_2 = 1 $ and $\gamma_1 \leq \gamma_2$. At each iteration, we pick the top element in the priority queue to generate the corresponding cylinder, and repeat the steps. Essentially, Equation~\ref{eqn:algo_1} selects a grid point which is furthest from the current point and is going in a similar direction. This approach tackles noise/errors in local orientation better than integral curves by looking at multiple choices for vertices and avoiding intra cell interpolation in an already noisy environment.

In the example in Figure~\ref{fig:algo}b we select the orange grid vertex next and repeat the process. We do not select the purple vertex (Figure~\ref{fig:algo}c) because it is not a candidate vertex. If we generate tracts that have erroneous local orientations they will not be able to find further possible candidate vertices and will be of short length. Short tracts (less than 0.01 times the smallest axes size) are removed. The MetaTracts generated are shown in Figure~\ref{fig:length_distribution}a. The MetaTracts are colored with the mean orientation direction mapped to the RGB space. Figure~\ref{fig:length_distribution}a shows a single slice of YZ plane. Consistent orientation is a key intrinsic feature in the data becomes visually pronounced.
We apply uniform, dense seeding to the XCT volume data to trace and generate the tracts of fiber bundles. Figure~\ref{fig:length_distribution}b shows the MetaTracts of a particular bundle colored according to their individual length.

@@

The output of the above step is a set of MetaTracts. The next step is to cluster the MetaTracts into bundles. 
Carbon fibers have both orientation and geometric proximity information, both of which can be used for clustering. Experimentally we found that the orientation measure was very reliable, while measures of geometric proximity had problems with fibers which partially overlap (Figure~\ref{fig:length_distribution}b).  
Instead of creating a heuristic which artificially combines the orientation and geometric proximity measures, we first cluster based on orientation and then further subdivide each orientation cluster using geometric proximity.
We found that different clustering techniques performed preferably for different measures.  The two step clustering also made the problem more tractable and  the parameters intuitive for the end users. For the orientation clustering, we use dimension reduction followed by K-means clustering. We then use hierarchical clustering to further subdivide each cluster based on the geometric proximity measures.

@@

\subsection {Orientation based clustering}
%We use the inherent features expressed as weaving pattern of our data to pre-cluster the metaTracts into the number of classes based on the major directions of the pattern.
%It is important to note that this step broadly divides the MetaTracts into classes based on the major orientations and not individual fiber bundles. 
This step divides the individual MetaTracts into classes based on their major orientations. In order to cluster MetaTracts going in same directions, we use a spectral embedding technique called Laplacian eigenmaps which was originally introduced by Belkin and Niyogi \cite{Belkin01}. An eigenvalue problem is solved, which maps the manifold embedded in the graph into a lower dimensional space while preserving the graph structure.
Let $G$ be the graph, we compute the eigenvalues and eigenvectors for the generalized eigenvector problem $L\textit{\textbf{f}}=\lambda D\textit{\textbf{f}}$,
%\begin{equation}\label{equn:eigenMaps}
%L\textbf{f}=\lambda D\textbf{f}
%\end{equation}
where $D$ is the diagonal weight matrix and $L$ is the Laplacian matrix. The eigenvector \textbf{${f}_{0}$} corresponding to the eigenvalue 0 is left out and the next $m$, {\textbf{${f}_{1}$} through \textbf{${f}_{m}$}} eigenvectors are used to embed in an $m$-dimensional space (see section~\ref{sec:param_choices} for values of m).
 In our case, each MetaTract is a data point. We introduce a simple \textit{orientation based similarity measure}. There is no spatial information involved, just a partition of the MetaTracts based on orientation which is already inherent in the data.
Following this, given a pair of  MetaTracts, we define the edge weights between two tracts as the cosine of the maximum angle between the local orientations ($N_P$) of all pairs of start points ($C_P$) between the two MetaTracts. The edge weights give a distance matrix representing the distance between each pair of points. Using the Belkin and Niyogi algorithm we ``embed" these points in a low dimensional space where the Euclidean distance between points approximates the distance between points given by the distance matrix. 
%
%August 13 2014
%For all our tests the fibers were then embedded in a $\eta$ low dimensional space. The number of dimensions in the high dimensional space is equal to the number of datapoints or the number of MetaTracts in our case.
%For all our test cases we set $\eta$ to be five (see sec~\ref{sec:param_choices} for parameter choices for $\eta$.).
%  
We used conventional K-means for clustering this lower dimensional space. Here $K$ is supplied by domain knowledge of the number of major fiber bundle directions of the woven structure.
For our test case there are two major directions of the fiber bundles and $K$ was set to 2. Due to the advantages offered by the dimensionality reduction, even if there are curved fibers bundles the user has to provide just the major fiber bundle directions of the weaving pattern. Figure~\ref{fig:orientation_clustering}b shows the result of the K-means clustering with the data points projected to the top three eigenvectors as the major axes. As is expected there is a clear distinction based on fiber orientation. Figure~\ref{fig:orientation_clustering}a,c shows the MetaTracts colored according to clustering results.

@@

Unfortunately, the Hausdorff distance does not work well for our application since a fiber bundle may have many MetaTracts, which only partially cover the bundle (Figure~\ref{fig:length_distribution}b). If a MetaTract $P$ covers only part of the fiber bundle covered by $Q$, then $H_{dir}(P,Q)$ will be very small while $H_{dir}(Q,P)$ will be large.

@@

\subsection{Choice of clustering techniques}
As opposed to the two phase clustering we experimented with using proximity alone and using single step clustering of MetaTracts.
%\begin{figure}[tb]
%\centering
%		\includegraphics[width=0.4\textwidth]{images_pvis/figure7}
%		%\includegraphics[width=0.5\textwidth, trim = 0mm 70mm 0mm 20mm, clip]{imagesMT2014/image7b.pdf}
%	%\includegraphics[width=0.45\textwidth]{imagesMT2014/crop-16/bundles}~ 
%	%\includegraphics[height=0.35\textwidth]{imagesMT2014/crop-16/vol_mesh.png} 
%	\caption{(A) Volume rendering of data set Fig.~\ref{fig:data-char}A.(B) shows a single slice of the data set. (C) shows two of the extracted meshes together. (D) shows the meshes separately.}
%	\label{fig:crop-16-decomp}
%\end{figure}
